String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-08-21 2023-1-19/sections/introduction.tex
some updates

In modern programming languages, such as JavaScript, Python, Java, and PHP, the string data type plays a crucial role. 
String manipulations are error-prone and could even give rise to severe security vulnerabilities (e.g. \ cross-site scripting, aka XSS). 
One powerful method for identifying such bugs is \emph{symbolic execution} (possibly in combination with dynamic analysis). It analyses symbolic paths in a program by viewing them as constraints checked by constraint solvers. 
%
Symbolic execution of string manipulating programs has motivated the highly active research area of \emph{string constraint solving}, resulting in the development of numerous string solvers in the last decade, e.g.,
Z3seq~\cite{z3seq}, CVC4/5~\cite{cvc4}, Z3str/2/3/4~\cite{Z3-str,Z3-str2,Z3-str3,BerzishMurphy2021}, Z3str3RE~\cite{BD+23}, 
Z3-Trau~\cite{Z3-trau}, OSTRICH~\cite{CHL+19}, Slent~\cite{WC+18}, among many others.

@@

Nevertheless, the aforementioned state-of-the-art string constraint solvers still suffer from such string constraints, especially when the counting bounds are large. For instance, none of the string solvers CVC5, Z3seq, Z3-Trau, Z3str3, Z3str3RE, and OSTRICH is capable of solving the following constraint within 300 seconds, and Z3 can not solve it within 120 seconds.
%
\begin{equation}\label{eqn-running}
x \in (\Sigma \setminus a)^{\{1, 60\}} (\Sigma \setminus b)^{\{1, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c^+ \wedge |x| > 120.
\end{equation}
Intuitively, the constraint in (\ref{eqn-running}) specifies that $x$ is a concatenation of three strings $x_1, x_2, x_3$ where $a$ (resp. $b, c$) does not occur in $x_1$ (resp. $x_2, x_3$), moreover, $x$ ends with a nonempty sequence of $c$'s, and the length of $x$ is greater than $120$. It is easy to observe that this constraint is unsatisfiable since on the one hand, $|x| > 120$ and the counting upper bound $60$ in both $(\Sigma \setminus a)^{\{1, 60\}}$ and $(\Sigma \setminus b)^{\{1, 60\}}$ imply that $x$ must end with some character from $\Sigma \setminus c$, that is, a character different from $c$, and on the other hand, $x \in \Sigma^*c^+$ requires that $x$ has to end with $c$.

A typical way for string constraint solvers to deal with regular expressions with counting is to unfold them into those \emph{without} counting using the concatenation operator. For instance, $a^{\{1, 4\}}$ is unfolded into $a(\varepsilon + a + aa + aaa)$ and $a^{\{2,\infty\}}$ is unfolded into $aaa^{*}$. Since the unfolding incurs an exponential blow-up on the sizes of constraints (assuming that the counting in string constraints are encoded in binary), the unfolding penalizes the performance of the solvers especially when the bounds are large.

@@

\medskip
\noindent 
\emph{Contribution.} In this work, we focus on the class of string constraints with regular membership and string-length function, where the counting operators may occur (called RECL for brevity). We make the following contributions: 
\vspace{-0.5mm}
\begin{itemize}
  \item We propose an automata-theoretical approach for solving RECL constraints. 
  Our main idea is to represent the counting operators by registers in cost-enriched finite automata (CEFA, see Section~\ref{sec:automaton} for the definition), instead of unfolding them explicitly. The satisfiability of RECL constraints is reduced to the nonemptiness problem of CEFA w.r.t. a linear integer arithmetic (LIA), which is then transformed to the satisfiability of existential LIA formula tackled by the off-the-shelf SMT solver.
  % Moreover, the string-length function can also be modeled by register in CEFA. Then the satisfiability of RECL constraints is reduced to the nonemptiness problem of CEFA with respect to a linear integer arithmetic (LIA) formula whose free variables are registers. According to the results from~\cite{atva2020}, an LIA formula can be computed to represent the potential values of registers in CEFA. 
  % Thus, the nonemptiness of CEFA w.r.t. LIA formulas can be reduced to the satisfiability of LIA formulas, which is then tackled by off-the-shelf SMT solvers.
  \item We propose size-reduction techniques to reduce the state space and the number of transitions in CEFA, to achieve better performance. It is beneficial since the sizes of LIA formulas depend much on the sizes of CEFA.
  \item Combined with the size-reduction techniques mentioned above, the register representation of regex-counting and string-length in CEFA entails an efficient algorithm for solving RECL constraints. We implement the algorithm on top of OSTRICH, resulting in a practical string solver called $\ostrichrecl$. 
  \item Finally, we utilize a collection of benchmark suites comprising 48,843 instances in total to evaluate the performance of our tool. The result shows that $\ostrichrecl$ solves the most number of instances among all the solvers, with comparable speed (See Table~\ref{tab:results_regcol}). In addition, we carry out some detailed experiments, including evaluation of size-reduction techniques  and the experiment of large-counting benchmark. The result shows that the size-reduction techniques are effective in practice (See Table~\ref{tab:results_simp}), and $\ostrichrecl$ performance much better than other solvers on large-counting benchmark (See Fig~\ref{fig:table_large_count}).
  % The experiment results show that $\ostrichrecl$ can solve 48,698 out of 48,843 instances, while Z3str3RE, the state-of-the-art solver on RECL constraints, solves only 47,607 instances (where the timeout period is set as 60 seconds). Moreover, the speed of $\ostrichrecl$ is comparable to that of Z3str3RE. In comparison, CVC5, Z3seq, Z3-Trau, and Z3str3 solve only 44,704, 46,426, 24,121, and 35,866 instances respectively. (See Table~\ref{tab:results_regcol}-\ref{tab:results_summary} for more details.)
\end{itemize}
\vspace{-0.5mm}
% We propose an automata-theoretical approach for solving RECL constraints. Our main idea is to represent the counting operators by registers in automata (as much as possible) instead of unfolding them explicitly. Our approach is built upon cost-enriched finite automata (CEFA), proposed in~\cite{atva2020} to solve string constraints involving the integer data type. Moreover, the string-length function can also be modeled by register in CEFA. Then the satisfiability of RECL constraints is reduced to the nonemptiness problem of CEFA with respect to a linear integer arithmetic (LIA) formula whose free variables are registers of CEFA.

>>>>

The string data type plays a crucial role in modern programming languages, such as JavaScript, Python, Java, and PHP. 
String manipulations are error-prone and could even give rise to severe security vulnerabilities (e.g. \ cross-site scripting, aka XSS). 
One powerful method for identifying such bugs is \emph{symbolic execution} (possibly in combination with dynamic analysis). It analyses symbolic paths in a program by viewing them as constraints checked by constraint solvers. 
%
Symbolic execution of string manipulating programs has motivated the highly active research area of \emph{string constraint solving}, resulting in the development of numerous string solvers in the last decade, e.g.,
Z3seq~\cite{z3seq}, CVC4/5~\cite{cvc4}, Z3str/2/3/4~\cite{Z3-str,Z3-str2,Z3-str3,BerzishMurphy2021}, Z3str3RE~\cite{BD+23}, 
Z3-Trau~\cite{Z3-trau}, OSTRICH~\cite{CHL+19}, Slent~\cite{WC+18}, among many others.

@@

Nevertheless, the aforementioned state-of-the-art string constraint solvers still suffer from such string constraints, especially when the counting and length bounds are large. For instance, none of the string solvers CVC5, Z3seq, Z3-Trau, Z3str3, Z3str3RE, and OSTRICH is capable of solving the following constraint within 300 seconds, and Z3 can not solve it within 120 seconds.
%
\begin{equation}\label{eqn-running}
x \in (\Sigma \setminus a)^{\{1, 60\}} (\Sigma \setminus b)^{\{1, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c^+ \wedge |x| > 120.
\end{equation}
Intuitively, the constraint in (\ref{eqn-running}) specifies that $x$ is a concatenation of three strings $x_1, x_2, x_3$ where $a$ (resp. $b, c$) does not occur in $x_1$ (resp. $x_2, x_3$), moreover, $x$ ends with a nonempty sequence of $c$'s, and the length of $x$ is greater than $120$. It is easy to observe that this constraint is unsatisfiable since on the one hand, $|x| > 120$ and the counting upper bound $60$ in both $(\Sigma \setminus a)^{\{1, 60\}}$ and $(\Sigma \setminus b)^{\{1, 60\}}$ imply that $x$ must end with some character from $\Sigma \setminus c$, that is, a character different from $c$, and on the other hand, $x \in \Sigma^*c^+$ requires that $x$ has to end with $c$.

A typical way for string constraint solvers to deal with regular expressions with counting is to unfold them into those \emph{without} counting using the concatenation operator. For instance, $a^{\{1, 4\}}$ is unfolded into $a(\varepsilon + a + aa + aaa)$ and $a^{\{2,\infty\}}$ is unfolded into $aaa^{*}$. Since the unfolding incurs an exponential blow-up on the sizes of constraints (assuming that the counting in string constraints are encoded in binary), the unfolding penalizes the performance of the solvers, especially when the bounds are large.

@@

\medskip
\noindent 
\emph{Contribution.} In this work, we focus on the class of string constraints with regular membership and string-length function, where the counting operators may occur (called RECL for brevity). We make the following contributions: 
\vspace{-0.5mm}
\begin{itemize}
  \item We propose an automata-theoretical approach for solving RECL constraints. 
  Our main idea is to represent the counting operators by registers in cost-enriched finite automata (CEFA, see Section~\ref{sec:automaton} for the definition), instead of unfolding them explicitly. The satisfiability of RECL constraints is reduced to the nonemptiness problem of CEFA w.r.t. a linear integer arithmetic (LIA), which is then transformed to the satisfiability of existential LIA formula tackled by the off-the-shelf SMT solver.
  % Moreover, the string-length function can also be modeled by register in CEFA. Then the satisfiability of RECL constraints is reduced to the nonemptiness problem of CEFA with respect to a linear integer arithmetic (LIA) formula whose free variables are registers. According to the results from~\cite{atva2020}, an LIA formula can be computed to represent the potential values of registers in CEFA. 
  % Thus, the nonemptiness of CEFA w.r.t. LIA formulas can be reduced to the satisfiability of LIA formulas, which is then tackled by off-the-shelf SMT solvers.
  \item We propose size-reduction techniques to reduce the state space and the number of transitions in CEFA, to achieve better performance. It is beneficial since the sizes of LIA formulas depend much on the sizes of CEFA.
  \item Combined with the size-reduction techniques mentioned above, the register representation of regex-counting and string-length in CEFA entails an efficient algorithm for solving RECL constraints. We implement the algorithm on top of OSTRICH, resulting in a practical string solver called $\ostrichrecl$. 
  \item Finally, we utilize a collection of benchmark suites comprising 48,843 instances in total to evaluate the performance of our tool. The result shows that $\ostrichrecl$ solves the most instances among all the solvers, with comparable speed (See Table~\ref{tab:results_regcol}). In addition, we carry out some detailed experiments, including the evaluation of size-reduction techniques and the experiment of large-counting benchmarks. The result shows that the size-reduction techniques are effective in practice (See Table~\ref{tab:results_simp}), and $\ostrichrecl$ performs much better than other solvers on the large-counting benchmark (See Fig~\ref{fig:table_large_count}).
  % The experiment results show that $\ostrichrecl$ can solve 48,698 out of 48,843 instances, while Z3str3RE, the state-of-the-art solver on RECL constraints, solves only 47,607 instances (where the timeout period is set as 60 seconds). Moreover, the speed of $\ostrichrecl$ is comparable to that of Z3str3RE. In comparison, CVC5, Z3seq, Z3-Trau, and Z3str3 solve only 44,704, 46,426, 24,121, and 35,866 instances respectively. (See Table~\ref{tab:results_regcol}-\ref{tab:results_summary} for more details.)
\end{itemize}
\vspace{-0.5mm}
% We propose an automata-theoretical approach for solving RECL constraints. Our main idea is to represent the counting operators by registers in automata (as much as possible) instead of unfolding them explicitly. Our approach is built upon cost-enriched finite automata (CEFA), proposed in~\cite{atva2020} to solve string constraints involving the integer data type. Moreover, the string-length function can also be modeled by register in CEFA. Then the satisfiability of RECL constraints is reduced to the nonemptiness problem of CEFA with respect to a linear integer arithmetic (LIA) formula whose free variables are registers of CEFA.
